---
title: "Migration Dynamics and Quality of Life Index"

format:
  html:
    code-fold: true
    code-summary: "Code"
    toc: true             
    toc-location: left  
    theme: cosmo        
    embed-resources: true
  pdf:
    colorlinks: true
    papersize: a4
    fontsize: 11pt
    margin-left: 2cm
    margin-right: 2cm
    margin-top: 2cm
    margin-bottom: 2cm
    
execute:
  warning: false
  message: false
---

Welcome to our project page.

*Team BU-R-SEL*

[Selin DÜZEN](https://emu-hacettepe-analytics.github.io/emu660-spring2026-Selindznn/project.html)

[Burcu BARIŞ](https://emu-hacettepe-analytics.github.io/emu660-spring2026-burcubaris/project.html)

# 1. Project Overview and Scope

Migration is not merely the geographical displacement of individuals; it is also a reflection of economic pursuits, social opportunities, and structural quality in a region. In Turkey, the difference in development and living standards between provinces constitutes one of the primary dynamics of internal migration movements.

In this project, using TUIK (Turkish Statistical Institute) data, we aim to examine the multidimensional relationship between Net Migration (Net Migration = Migrant Receiving - Migrant Giving) figures for all 81 provinces of Turkiye and the Livability Index indicators in these provinces using analytical methods. Our goal is to reveal, through data analytics and statistical models, which vital factors (economy, education, health, environmental quality, security, etc.) cause people to leave their current cities or move to specific cities.

*Creating Migration Profiles:* To identify the "Net Migration Champions" (provinces receiving the most net migrants) and "Migration Sources" (provinces giving the most net migrants) in Turkiye and to map them geographically.

*The Impact of Livability Dimensions:* To statistically demonstrate which of the sub-components of the livability index (Income and Wealth, Health, Education, Environment, Security, etc.) are more dominant "push" or "pull" forces influencing migration decisions.

# 2. Data

## 2.1 Data Source

The datasets used in the project are:

1.[Population size of the provinces](https://veriportali.tuik.gov.tr/api/tr/data/downloads?t=i&p=btt%2FI0Npd%2BLvtOvCSZRIUZxbvXE3lIAf5bE6X3CdLQJxPknvezk%2BE6i3Md3GTPYHYMhzGdG2vbYkzOXomic6rY8F%2FyL8NGud1ChYCxoEgy5FkIT%2BMezMLoMVyKJUHwSgAQ%2BDeuZaUgEdXmyanOwSEQ%3D%3D)

2.[Migration received by provinces according to the reason for migration](https://veriportali.tuik.gov.tr/api/tr/data/downloads?t=i&p=%2FMyiBh3DmS%2Bv%2BBjCf9%2BzPq%2B96lw8HZTF50uTl9jqBks%2BK5WE6sQNt%2Fqka0J5%2BImWGTGxKxiPcrz7kNG6I3rKkrLi%2B8mZkMkz9reke8%2BHd%2Fo%3D)

3.[Migration rates by provinces according to the reason for migration](https://veriportali.tuik.gov.tr/api/tr/data/downloads?t=i&p=%2FMyiBh3DmS%2Bv%2BBjCf9%2BzPq%2B96lw8HZTF50uTl9jqBks%2BK5WE6sQNt%2Fqka0J5%2BImWGTGxKxiPcrz7kNG6I3rKkrLi%2B8mZkMkz9reke8%2BHd%2Fo%3D)

4.\[Net migration rate for provinces according to reasons for migration\] Calculated as "Migration Outgoing - Migration Incoming"

5.[Life Index Rankings and Index Values by Province](https://veriportali.tuik.gov.tr/api/tr/data/downloads?t=i&p=5IS1p1Z%2Btvlc0qxFhZzbC9g6NOWXugwdILs8mAMegx5w5sIpUqxydG6%2B5%2FWn1HGIIXJVqd%2Fwn%2B78duD2fsK84VEREA5Sn4QY5oeOWRHZRvVYOaDudiOCwll9hFpmp5W)

In this study, the five datasets mentioned above were used, graphs were plotted, and the results were thoroughly analyzed.

## 2.2 General Information About Data

Our dataset tracks migration metrics across all 81 provinces of Turkiye from 2018 to 2024. It combines raw inbound/outbound migration figures with detailed columns explaining 12 different reasons for migration, cross-references them with the Turkish Statistical Institute's (TUIK) comprehensive Life Index scores, and shows the relationships between them.

## 2.3 Reason of Choice

We chose this data because migration cannot be explained solely by economic factors. By combining net migration metrics with the multidimensional Life Index (which assesses health, education, security, and environment), we can reveal the relationship between internal migration in Turkiye and life indices.

## 2.4 Preprocessing

The net migration values are calculated using migration data received and given by provinces in different Excel spreadsheets and the char formatted values containing spaces in the data were updated to numeric values.

```{r}
#install.packages("ggplot2") 
#install.packages("dplyr")   
#install.packages("openxlsx")
#install.packages("readxl")

# 1. R'ın geçici işlemler için C diskinin altında temiz bir klasör kullanmasını sağlıyoruz
write("TMPDIR = C:\\", file = file.path(Sys.getenv("HOME"), ".Renviron"), append = TRUE)

# 2. Şimdi paketi tekrar yüklemeyi dene
#install.packages("writexl")
#install.packages("ggrepel")
library(ggplot2)
library(dplyr)
library(readxl) 
library(writexl) 
library(stringr)
library(forcats)
library(tidyr)
library(openxlsx)
library(ggplot2)
library(readxl)
library(dplyr)
library(ggrepel)
```

### 2.4.1. Calculating net migration values and creating a separate Excel file and checking and correcting numerical values.

```{r}
VerilenGoc <- read_excel("C:/emu660-spring2026-burcubaris/VerilenGoc.xlsx", col_names = TRUE)
AlinanGoc  <- read_excel("C:/emu660-spring2026-burcubaris/AlinanGoc.xlsx", col_names = TRUE)
toplam_sutun <- ncol(VerilenGoc)
fark_df <- VerilenGoc 
for (i in 3:toplam_sutun) {
  v_clean <- as.character(VerilenGoc[[i]])
  v_clean <- str_replace_all(v_clean, "\\s", "") 
  VerilenGoc[[i]] <- as.numeric(v_clean)        
  a_clean <- as.character(AlinanGoc[[i]])
  a_clean <- str_replace_all(a_clean, "\\s", "")    
  AlinanGoc[[i]] <- as.numeric(a_clean)
}
fark_df[, 3:toplam_sutun] <- VerilenGoc[, 3:toplam_sutun] - AlinanGoc[, 3:toplam_sutun]
for (i in 3:toplam_sutun) {
  fark_df[[i]] <- as.numeric(fark_df[[i]])
}
write_xlsx(fark_df, "C:/emu660-spring2026-burcubaris/NetGoc.xlsx")
rmarkdown::paged_table(VerilenGoc[, 3:toplam_sutun])
```
Bu tablodan illerin en büyük göç verme nedeninin eğitim olduğunu görebiliyoruz. Eğitimi ekonomik/iş istihdam olanakları takip etmektedir.

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# 1. Veriyi hazırla ve filtrele (Sütun ismi yerine indeks numarası kullanıyoruz)
neden_verisi_2024 <- VerilenGoc %>%
  # 1. sütun Yıl (2024) ve 2. sütun İl (Toplam-Total) olan satırı seçiyoruz
  filter(.[[1]] == 2024 & .[[2]] == "Toplam-Total") %>%
  # İlk 3 sabit sütunu (Yıl, İl, Toplam) indeks numaralarıyla tablodan çıkarıyoruz
  select(-c(1, 2, 3)) %>% 
  # Tabloyu ggplot'un anlayacağı uzun (long) formata çeviriyoruz
  pivot_longer(cols = everything(), names_to = "Goc_Nedeni", values_to = "Kisi_Sayisi") %>%
  mutate(Kisi_Sayisi_Bin = Kisi_Sayisi / 1000)

# Sütun isimlerindeki o sorun çıkaran \r\n karakterlerini temizliyoruz
neden_verisi_2024$Goc_Nedeni <- gsub("\\r\\n.*", "", neden_verisi_2024$Goc_Nedeni) 

# 2. Grafiği Çiz (İller grafiğindeki sıkışma olmasın diye yazı boyutunu da ayarladık)
ggplot(neden_verisi_2024, aes(x = Kisi_Sayisi_Bin, y = reorder(Goc_Nedeni, Kisi_Sayisi_Bin))) +
  geom_col(fill = "forestgreen", alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "2024 Yili Goc Nedenleri",
    x = "Kisi Sayısı (Bin Kisi)",
    y = "Goc Etme Gerekcesi"
  ) +
  theme(
    axis.text.y = element_text(size = 9), # Gerekçelerin yazı boyutu
    plot.title = element_text(face = "bold", size = 12),
    panel.grid.major.y = element_blank() # Yatay kalabalık çizgileri gizler
  )
```  
  
En güncel 2024 baz alırsak göç etmenin en büyük nedenlerinin aile fertlerinden birine bağımlı göç ile daha iyi yaşam koşulları ve eğitim olduğunu görürüz. 2024 yılı için en yüksek çıkan değer aslında bize Türkiye'de yaşayan toplumun aile odaklı olduğunu gösterir ki bu sürpriz değildir.İkinci olarak ise insanların daha iyi yaşam koşulları için yoğun bir şekilde yer değiştirdiğini görüyoruz.Üçüncü sırada çıkan eğitimde ise üniversite ve okul faktörünün yine Türkiye'de iç göçün temellerinden olmaya devam ettiğini görüyoruz.


### 2.4.2. Calculating net migration rate using population data of provinces and creating an Excel file.

```{r}
kaynak_yolu <- "C:/emu660-spring2026-burcubaris/NetGoc.xlsx"
hedef_yolu  <- "C:/emu660-spring2026-burcubaris/IlNufuslarii.xlsx"

df_kaynak <- read_excel(kaynak_yolu, col_names = TRUE)

NetGocVerisi <- df_kaynak[1:82, 3, drop = TRUE] 
wb <- loadWorkbook(hedef_yolu)
writeData(wb, sheet = 1, x = "Net GC6C'", startCol = 3, startRow = 1, colNames = FALSE)
writeData(wb, sheet = 1, x = NetGocVerisi, startCol = 3, startRow = 2, colNames = FALSE)

saveWorkbook(wb, hedef_yolu, overwrite = TRUE)
```

# 3. Analysis

## 3.1 Exploratory Data Analysis

### 3.1.1. Changes in Total Migration Values by Year (2018-2024)

The changes in the total amount of migration over the years were shown graphically.

```{r}
grafik_verisi <- VerilenGoc[VerilenGoc[, 2] == "Toplam-Total", ]
print(grafik_verisi)
head(grafik_verisi)

yil <- factor(grafik_verisi[[1]], levels = sort(unique(grafik_verisi[[1]]), decreasing = TRUE))
goc_miktari  <- as.numeric(grafik_verisi[[3]])/1000  

ggplot(data = grafik_verisi, aes(x = yil, y = goc_miktari)) +
  geom_col(fill = "steelblue", alpha = 0.7) + 
  geom_line(color = "black", size = 2) + 
  geom_point(color = "darkred", size = 3) +
  geom_text(aes(label = round(goc_miktari, 1)), vjust = -0.5, size = 3.5) +
  labs(
    title = "Yillara Gore Toplam Verilen Goc Miktari",
    x = "Yil",
    y = "Goc Miktari (x1.000)"
  )
```

2018-2024 yılları arasındaki göç hareketini inceleyecek olursak en düşük yılın 2020 pandemi senesinin olduğunu görüyoruz. O dönemde sokağa çıkma yasakları ve belirsizliğin hakim olması insanların yaşadıkları şehirden göç etmelerine engel olmuş olabilir diyebiliriz. En yüksek artışın ise 2023 yılında olduğunu görüyoruz ki Şubat 2023'te Kahramanmaraş depremi meydana gelmiş ve insanlar deprem riskinin az olduğu, daha güvenli şehirlere göç etmeye başlamışlardır. 2018 yılında da nispeten yüksek bir göç hareketi görüyoruz. O dönemde Türkiye'de ekonomik kriz ve enflasyon etkisiyle özellikle pahalı olan Ankara ve İstanbul gibi metropol şehirlerden daha küçük şehirlere göç hareketinin olduğunu söyleyebiliriz.

### 3.1.2. Changes in Total Migration Values by Year for Ankara Provience

The graphic shows the changes in total migration amounts over the years for Ankara province.

```{r}
library(ggplot2)

# 1. Ankara verisini filtreliyoruz
grafik_verisi <- VerilenGoc[VerilenGoc[, 2] == "Ankara", ]

# 2. Yeni degiskenleri tablonun disinda olusturmak yerine, 
# dogrudan tabloya yeni sutun olarak ekliyoruz (ggplot'un en sevdigi yontem)
grafik_verisi$yil_farkli <- factor(grafik_verisi[[1]], levels = sort(unique(grafik_verisi[[1]]), decreasing = TRUE))
grafik_verisi$goc_bin <- as.numeric(grafik_verisi[[3]]) / 1000

# 3. Grafigi cizdiriyoruz (Artik eksenler tablonun icindeki yeni sutunlari goruyor)
ggplot(data = grafik_verisi, aes(x = yil_farkli, y = goc_bin, group = 1)) +
  geom_col(fill = "steelblue", alpha = 0.7) + 
  geom_line(color = "black", linewidth = 1.5) + # size yerine yenilenen linewidth kullanildi
  geom_point(color = "darkred", size = 3) +
  geom_text(aes(label = round(goc_bin, 1)), vjust = -0.5, size = 3.5) +
  theme_minimal() + # Daha temiz bir gorunum ekledik
  labs(
    title = "Ankara Icin Yillara Gore Toplam Verilen Goc Miktari",
    x = "Yil",
    y = "Goc Miktari (x1.000)"
  )
```

### 3.1.3. Migration Figures by Province in 2024

```{r}
#| fig-height: 12
#| fig-width: 8
AlinanGoc  <- read_excel("C:/emu660-spring2026-burcubaris/AlinanGoc.xlsx",col_names = TRUE)

colnames(AlinanGoc)[1:3] <- c("Yil", "Il", "Goc_Miktari")
veri_2024 <- subset(AlinanGoc, Yil == 2024 & Il != "Toplam-Total")
veri_2024$Goc_Miktari_Bin <- veri_2024$Goc_Miktari / 1000


ggplot(veri_2024, aes(x = Goc_Miktari_Bin, y = reorder(Il, Goc_Miktari_Bin))) +
  geom_col(aes(fill = Goc_Miktari_Bin > 0)) +
  scale_fill_manual(values = c("coral", "skyblue"), guide = "none") +
  theme_minimal() +
  labs(
    title = "2024 Yilinda Illerin Aldigi Goc Miktarlari",
    x = "Alinan Goc Miktari (Bin Kisi)",
    y = "Iller"
  )
```

### 3.1.4. Net Migration Figures of Provinces in 2024

A graphical representation of the calculation made using the previously calculated data given migration-received migration data.

```{r}
NetGoc  <- read_excel("C:/emu660-spring2026-burcubaris/NetGoc.xlsx",col_names = TRUE)

colnames(NetGoc)[1:3] <- c("Yil", "Il", "Goc_Miktari")
veri_2024 <- subset(NetGoc, Yil == 2024 & Il != "Toplam-Total")
veri_2024$Goc_Miktari_Bin <- veri_2024$Goc_Miktari / 1000


ggplot(veri_2024, aes(x = Goc_Miktari_Bin, y = reorder(Il, Goc_Miktari_Bin))) +
  geom_col(aes(fill = Goc_Miktari_Bin > 0)) +
  scale_fill_manual(values = c("coral", "skyblue"), guide = "none") +
  theme_minimal() +
  labs(
    title = "2024 Yilinda Illerin Net Goc Miktarlari",
    x = "Net Goc Miktari (Bin Kisi)",
    y = "Iller"
  )  
```

### 3.1.5. Graph showing the distribution of migration reasons that make up net migration data.

Migration reasons include: change of job, starting/finding a job, change of marital status/family-related reasons, better housing and living conditions, migration related to any member of the household/family, returning to family home/hometown, health/care, buying a house, retirement, other, and unknown.

```{r}
NetGoc  <- read_excel("C:/emu660-spring2026-burcubaris/NetGoc.xlsx",col_names = TRUE)

goc_nedenleri <- c("Tayin/Is Degisikligi", "Ise Baslamak/Is Bulmak", "Egitim", 
                   "Medeni Durum/Ailevi", "Daha Iyi Konut/Yasam", "Haneye Bagimli Goc", 
                   "Memlekete Donus", "Saglik/Bakim", "Ev Alinmasi", 
                   "Emeklilik", "Diger", "Bilinmeyen")

colnames(NetGoc)[1:3] <- c("Yil", "Il", "Toplam_Goc")
colnames(NetGoc)[4:15] <- goc_nedenleri
veri_2024 <- subset(NetGoc, Yil == 2024 & Il != "Toplam-Total")

veri_uzun <- pivot_longer(veri_2024, 
                          cols = 4:15, 
                          names_to = "Goc_Nedeni", 
                          values_to = "Miktar")

veri_uzun$Miktar_Bin <- veri_uzun$Miktar / 1000

goc_grafigi <- ggplot(veri_uzun, aes(x = Miktar_Bin, y = reorder(Il, Toplam_Goc), fill = Goc_Nedeni)) +
  geom_col(position = "stack") +
  theme_minimal() +
  labs(
    title = "2024 Illerin Goc Nedenlerine Gore Dagilimi",
    subtitle = "Miktarlar bin kisi cinsinden ifade edilmistir",
    x = "Net Goc Miktari (Bin Kisi)",
    y = "Iller",
    fill = "Goc Nedenleri"
  ) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8),
    axis.text.y = element_text(size = 7)
  ) +
  guides(fill = guide_legend(ncol = 3))
  
print(goc_grafigi)
```

## 3.2 Trend Analysis

The graph shows the relational analysis between migration rate data and life indices of the provinces.

```{r}

yasam_df <- read_excel("C:/emu660-spring2026-burcubaris/YasamEndeksi.xlsx", col_names = TRUE)
goc_df <- read_excel("C:/emu660-spring2026-burcubaris/Gochizi.xlsx", col_names = TRUE)

colnames(yasam_df)[1:2] <- c("Il", "Endeks")
colnames(goc_df)[1:3]   <- c("Yil", "Il", "Goc")

goc_2024 <- subset(goc_df, Yil == 2024 & Il != "Toplam-Total")

veri_seti <- inner_join(yasam_df, goc_2024, by = "Il")

grafik <- ggplot(veri_seti, aes(x = Endeks, y = Goc, label = Il)) + 
  geom_point(color = "#1f77b4", size = 3, alpha = 0.7) + 
  geom_text_repel(size = 3, max.overlaps = 15) + 
  geom_smooth(method = "lm", color = "red", se = TRUE, formula = y ~ x) + 
  labs(
    title = "Il Yasam Endeksi ve 2024 Net Goc Hizi İliskisi",
    x = "Genel Yasam Endeksi",
    y = "Net Goc Hizi (‰)"
  ) +
  theme_minimal()

print(grafik)
```
## 3.3 Results

The graph shows the

# 4 Results and Key Takeaways

The graph shows the